Formal Verification of Real-Time Data Processing of the LHC Beam Loss Monitoring System: A Case Study

### 2011-03-24 fmics/paper.tex
Add description+definitions of model to section 4

To prove the equation above, we need to formalize all of its terms.
First, we make two items explicit: that the equation is to be understood as holding at each time step, and that the terms depend on the stream of input counts over time.
With these refinements, our statement becomes \[\forall{D\,n\,t}.\,\text{\textsf{RS} }D\,n\,t=\text{sum, given input stream $D$, for output $n$ at time $t$}\pm\text{acceptable error}\]

To formalize the term on the left hand side, we need to build a logical model of the SRS structure.
Our work proceeded in two phases, starting with a model of a simplified structure and proceeding to a more realistic model.
In both cases, we represent each piece of the SRS structure that has a value---for example, each cell in each shift register---as a function in HOL that takes the global input stream and current time step as input and returns the value of the piece of the structure at that time.
Our simplifications in the first phase include
\begin{enumerate}
\item Ignore the middle taps in each slice
\item A more regular arrangement of slices than the actual structure
\item Ignore subtractor and accumulator, by defining the output of a slice as the sum of its cells directly
\item Simplify timing <what? how?>
\item Use natural numbers, rather than finite words, for cell values
\end{enumerate}
In the second phase, we drop the first three simplifications by adding middle taps, a more flexible arrangement of slices, and the subtractor and accumulator to our model.
In both cases, however, the model is simultaneously a simplification and a generalization of the actual structure in the following sense: we model an infinite number of shift registers, each with an infinite number of cells, simply by not specifying their widths or any limits.
<how does that work>
<it's not completely clear that if we actually saved time/effort by doing the simplified one first...>

>>>>

To prove the equation above, we need to formalize all of its terms.
First, we make two items explicit: that the equation is to be understood as holding at each time step, and that the terms depend on the stream of input counts over time.
With these refinements, our statement becomes \[\forall{D\,n\,t}.\,\mathsf{RS}\;D\;n\;t=\text{sum, given input stream $D$, for output $n$ at time $t$}\pm\text{acceptable error}\]
(The input stream is a function of time: $D\;t$ is the input value to the whole SRS structure at time $t$.)

To formalize the term on the left hand side, we need to build a logical model of the SRS structure.
Our work proceeded in two phases, starting with a model of a simplified structure and proceeding to a more realistic model.
In both cases, we represent each piece of the SRS structure that has a value---for example, each cell in each shift register---as a function in HOL that takes the global input stream and current time step as input and returns the value of the piece of the structure at that time.
Our simplifications in the first phase include
\begin{enumerate}
\item Ignore the middle taps in each slice
\item A more regular arrangement of slices than the actual structure
\item Ignore subtractor and accumulator, by defining the output of a slice as the sum of its cells directly
\item Simplify timing <what? how?>
\item Use natural numbers, rather than finite words, for cell values
\end{enumerate}
In the second phase, we drop the first three simplifications by adding middle taps, a more flexible arrangement of slices, and the subtractor and accumulator to our model.
In both cases, however, the model is simultaneously a simplification and a generalization of the actual structure in the following sense: we model an infinite number of slices, each with an infinite number of shift register cells and taps, simply by letting indices range over the natural numbers without explicitly giving limits.

We describe the more realistic (second phase) model here.
Table~\ref{tab:descriptions} lists the names of the HOL functions we define and their intended meanings, and Figure~\ref{fig:definitions} gives their definitions. 
For our correctness equation above, we only need the last entry in the table, $\mathsf{RS}$; the first two entries, $\mathsf{tap}$ and $\mathsf{input}$, enable us to specify the exact configuration of the structure, and the remaining entries are used as ``glue'' leading up to the definition of $\mathsf{RS}$.
By using these intermediary functions, every definition in Figure~\ref{fig:definitions} is \emph{local} (only talks about a small part of the overall SRS structure) and therefore is easy to verify against its intended meaning.

\begin{table}
\caption{
Descriptions of the HOL functions comprising our model of the SRS structure.
\label{tab:descriptions}
}
\begin{tabular}{lp{0.8\textwidth}}
Function&Intended meaning\\
\hline\\
\(\mathsf{tap}\;n\;x\)&The position of tap $x$ of slice $n$.\\
\(\mathsf{input}\;n\)&A pair $(n',x)$ indicating that the input to slice $n$ is output $x$ of slice $n'$.\\
\(\mathsf{delay}\;n\)&The number of time steps between updates of slice $n$.\\
\(\mathsf{update\_time}\;n\;t\)&A boolean indicating whether $t$ is an update time for slice $n$.\\
\(\mathsf{source}\;D\;n\;m\;t\)&The value of the cell that is the direct input to cell $m$ of slice $n$, at time $t$, given input stream $D$.\\
\(\mathsf{SR}\;D\;n\;m\;t\)&The value of cell $m$ of shift register $n$, at time $t$, given input stream $D$.\\
\(\mathsf{output}\;D\;n\;x\;t\)&The value of output $x$ of slice $n$, at time $t$, given input stream $D$.\\
\(\mathsf{RS}\;D\;n\;t\)&The value of Running Sum register $n$ at time $t$, given input stream $D$.
\end{tabular}
\end{table}

\begin{figure}
\caption{
Definitions of the HOL functions comprising our model of the SRS structure.
\(\mathsf{tap}\) and \(\mathsf{input}\) are defined to match Figure~\ref{fig:srs}, counting positions within a shift register from $0$.
Slice $0$ is a fake slice representing the global input; this enables a succinct definition of $\mathsf{source}$.
The definition of $\mathsf{input}$ when $n=0$ or when $n>6$ does not change the structure represented, so we use definitions convenient for proving theorems.
A similar comment applies to other definitions when $n$, representing a slice number, is $0$, or when $x$, representing a shift register position, is above $1$.
\label{fig:definitions}}
\begin{align*}
\mathsf{tap}\;0\;0&=0&\mathsf{tap}\;0\;x&=0\\
\mathsf{tap}\;1\;0&=1-1&\mathsf{tap}\;1\;x&=2-1\\
\mathsf{tap}\;2\;0&=8-1&\mathsf{tap}\;2\;x&=16-1\\
&\dots&\mathsf{tap}\;6\;x&=128-1\quad\text{(where $x>0$)}\\
\mathsf{input}\;0&=(0,0)&\mathsf{input}\;1&=(0,0)\\
\mathsf{input}\;2&=(0,0)&\mathsf{input}\;3&=(1,1)\\
\mathsf{input}\;4&=(3,0)&\mathsf{input}\;5&=(4,0)\\
\mathsf{input}\;6&=(4,1)&\mathsf{input}\;n&=(n-1,0)\quad\text{(where $n>6$)}
\end{align*}
\begin{align*}
\mathsf{delay}\;0&=1\\
\mathsf{delay}\;n&=\mathsf{delay}\;n'\times((\mathsf{tap}\;n'\;x)+1)\\
&\qquad\text{where $(n',x)=\mathsf{input\;n}$ (and $n>0$)}
\end{align*}
\begin{align*}
\mathsf{update\_time}\;n\;t&\iff(t\operatorname{mod}\mathsf{delay}\;n=0)
\end{align*}
\begin{align*}
\mathsf{source}\;D\;n\;0\;t&=\mathsf{output}\;D\;n'\;x\;t\quad\text{where $(n',x)=\mathsf{input\;n}$}\\
\mathsf{source}\;D\;n\;m\;t&=\;\mathsf{SR}\;D\;n\;(m-1)\;\quad\text{(where $m>0$)}
\end{align*}
\begin{align*}
\mathsf{output}\;D\;0\;x\;t&=D\;t\\
\mathsf{output}\;D\;n\;x\;0&=0\\
\mathsf{output}\;D\;n\;x\;t&=
\text{if $\mathsf{update\_time}\;n\;t$}\\
&\qquad\text{then }((\mathsf{output}\;D\;n\;x\;(t-1))+(\mathsf{source}\;D\;n\;0\;(t-1)))-(\mathsf{SR}\;D\;n(\mathsf{tap}\;n\;x)\;(t-1))\\
&\qquad\text{else }\mathsf{output}\;D\;n\;x\;(t-1)\qquad\text{(where $t>0$)}
\end{align*}
\begin{align*}
\mathsf{SR}\;D\;n\;m\;0&=0\\
\mathsf{SR}\;D\;n\;m\;t&=\text{if $\mathsf{update\_time}\;n\;t$}\\
&\qquad\text{then }\mathsf{source}\;D\;n\;m\;(t-1)\\
&\qquad\text{else }\mathsf{SR}\;D\;n\;m\;t\qquad\text{(where $t>0$)}
\end{align*}
\begin{align*}
\mathsf{RS}\;D\;n\;t&=\mathsf{output}\;D\;\left(\left\lfloor\frac{n}{2}\right\rfloor+1\right)\;(n\operatorname{mod}2)
\end{align*}
\end{figure}
